HealpixNet: Efficient spherical Convolutional Neural Networks with Healpix sampling for cosmological applications

### 2018-06-20 scnn.tex
update simulation results
add new figures
lunch commit

\section{TODO list}
\begin{itemize}
    \item Michael, do you agree with the title HealpixNet
    \item Emphaise the novelty of using it on Helapix in abstract, into and rest of the paper
    \item Finish section on Convolutions on graphs (Michael - Nathanael)
    \item Plot of the new filters from Mass maps (Michael)
    \item (done!, plot to be added) check the power spectrum after smoothing
    \item (done!) upgrade the plot of pooling (Nathanael)
    \item new plot for network structure (teaser plot) (Michael)
    \item add info on comparison with histogram, so that it's just a demonstration
    \item (done!) Figure 1 to be round
    \item finish section on Efficient convolutions (Michael)
    \item conclusions
    \item (done!) new simulations with 3 arcmin smoothing
    \item (done!) simulation with cross-validation for the histogram
    \item (done!) Update result section according to the new simulations...
    \item (done!) simulation PSD
    \item (done!) Results figure, add sigma8
    \item possibly move the eigen decomposition and heat into appendix + add discussion (Nathanael)
\end{itemize}

@@

%\subsection{Motivation}
\assign{Tomek}

@@

The first approach, leveraging the continuous spherical convolution, has been
proposed by~\cite{cohen2017convolutional,cohen2018spherical}. The goal of these
contributions it to use a rotational equivariant convolution on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The resulting convolution is performed by a spherical Fourier transform (i.e. a projection
on the spherical harmonics), a  multiplication in the spectral domain and an
inverse spherical Fourier transform. Hence the main computational cost of a
convolution comes from the two Fourier transforms. Fortunately, the Fourier
transform can be accelerated for special samplings set (theoretically including
HealPix, \cite{...}). The main advantage of this approach is that it provides a
mathematically well defined rotational equivariant network. Nevertheless, even
with these Fourier transform acceleration, the convolutions remain expensive in
comparison to a traditional 2-dimensional convolution, limiting the practical use of this
approach. While a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the Healpix one. Hence, a practical comparison is
not possible.

@@

In our experiment we classify convergence maps, similar to those created with gravitational lensing technique \citep{chang2017curvedsky}.
We compare the performance of the spherical CNN to a simple benchmark algorithm.
This algorithm is based on classification of histogram of pixels in the maps.
We use an SVM to perform that task.
The benchmark method uses the information about the distribution of pixels, but does not include any spatial information.
The CNN can exploit the spatial information.
The performance is assessed by measuring the classification error on the test set.

@@

We make full sky N-body simulations for two parameter set for $\Lambda \rm{CDM} $ cosmological model: model 1 ($\Omega_m=0.31, \sigma_8=0.82$) and model 2 ($\Omega_m=0.26, \sigma_8=0.91$), where $\Omega_m$ is the matter density in the universe and $\sigma_8$ is the normalisation of the matter power spectrum.
Other parameter parameters were set to $h=0.7$ (MORE PARAMS HERE - CHECK WITH RAPHAEL).
These parameters were chosen to have the same spherical harmonic power spectrum for $\ell<1000$.
That means that it is not possible to distinguish between these cosmological models using these maps if the this range of scales is used.
We found that for $\ell>1000$ the differences in power spectrum is $~5\%$.
To remove this information, we additionally smooth the spherical maps with a Gaussian kernel.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.95\textwidth]{figures/smooth_map_class_1.pdf}
\includegraphics[width=0.95\textwidth]{figures/smooth_map_class_2.pdf}
\caption{Example maps from two classes. Top: model 1 ($\Omega_m=0.31, \sigma_8=0.82$). Bottom: model 2 ($\Omega_m=0.26, \sigma_8=0.91$).
The initial conditions for these simulations were the same, so differences arise only due to different cosmological parameters.}
\label{fig:map_sample}
\end{figure*}

@@

Leveraging the properties of the HealPix sampling, we can split each simulation
into $12*4^o$ ($o=0,1,2,\dots$) samples that span a smaller part of the sphere.
As shown in Figure~\ref{fig:subpart_sphere}, we used $o=0,1,2$ as the resulting
samples are large enough to suffer from the effects of the spherical geometry. We
decided to only report simulation results for this specific setting as spherical
cosmological data usually does not span the full sphere. Nevertheless, our code
includes an example using the full sphere.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/part_sphere.pdf}
\caption{3 subparts of the sphere with different sizes. Blue: $o=0$. Green: order $o=1$. Yellow: order $o=2$.}
\label{fig:subpart_sphere}
\end{figure}

We would like to build a robust classifier, i.e. an automatic way to
discriminate between the two classes with the presence of noise. For all the
experiments, we work with centered Gaussian noise. \todo{Tomek: could you put a
justification.} All classifiers are trained with noisy samples. To improve the
stability of the spherical CNN, we slowly increase the amount of noise added to
the training data over the iterations. Furthermore in order to avoid
overfitting, the noise is randomly produced during the training process. We trained our
model between $5$ and $40$ epochs depending on the complexity and the size of
the problem. The testing set was simply used to assess the global performance of
the network and to assess if it was overfitting.
Training took between $1$ and $10$ hours using a Nvidia 1080 TI.

We compare the spherical CNN with a handcrafted feature classifier. CITE \cite{...}
\todo{Tomek: we probably need 1 or two other classifiers.} The features are
constructed using a normalized histogram of the values. We then train a linear
SVN classifier. We tried different kernels, but they were not leading to better
results. For a fair comparison, we increase the amount of training data of the
SVN classifier until the testing error plateau to a fix value.

@@

>>>>

\section{TODO list}
\begin{itemize}
    \item Michael, do you agree with the title HealpixNet
    \item Emphaise the novelty of using it on Helapix in abstract, into and rest of the paper
    \item Finish section on Convolutions on graphs (Michael - Nathanael)
    \item Plot of the new filters from Mass maps (Michael)
    \item (done!, plot to be added) check the power spectrum after smoothing
    \item (done!) upgrade the plot of pooling (Nathanael)
    \item new plot for network structure (teaser plot) (Michael)
    \item add info on comparison with histogram, so that it's just a demonstration
    \item (done!) Figure 1 to be round
    \item finish section on Efficient convolutions (Michael)
    \item conclusions
    \item (done!) new simulations with 3 arcmin smoothing
    \item (done!) simulation with cross-validation for the histogram
    \item Update result section according to the new simulations...
    \item (done!) simulation PSD
    \item (done!) Results figure, add sigma8
    \item possibly move the eigen decomposition and heat into appendix + add discussion (Nathanael)
\end{itemize}

@@

%\subsection{Motivation}

@@

The first approach, leveraging the continuous spherical convolution, has been
proposed by~\cite{cohen2017convolutional,cohen2018spherical}. The goal of
these contributions it to use a rotational equivariant convolution on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The resulting convolution is performed by a spherical Fourier transform (i.e.
a projection on the spherical harmonics), a  multiplication in the spectral
domain and an inverse spherical Fourier transform. Hence the main
computational cost of a convolution comes from the two Fourier transforms.
Fortunately, the Fourier transform can be accelerated for special samplings
set (theoretically including HealPix, \cite{...}). The main advantage of this
approach is that it provides a mathematically well defined rotational
equivariant network. Nevertheless, even with these Fourier transform
acceleration, the convolutions remain expensive in comparison to a traditional
2-dimensional convolution, limiting the practical use of this approach. While
a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the Healpix one. Hence, a practical comparison is
not possible. We also note that this approach cannot be trivially accelerated
if only one part of the sphere is used. This is impractical in the field of
cosmology as the data often observed only on a subpart of the sphere.

@@

In our experiment we classify convergence maps, similar to those created with gravitational lensing technique \citep{chang2017curvedsky}.

@@

We make full sky N-body simulations for two parameter set for $\Lambda \rm{CDM} $ cosmological model: model 1 ($\Omega_m=0.31, \sigma_8=0.82$) and model 2 ($\Omega_m=0.26, \sigma_8=0.91$), where $\Omega_m$ is the matter density in the universe and $\sigma_8$ is the normalisation of the matter power spectrum.
Other parameter parameters were set to $h=0.7$ (MORE PARAMS HERE - CHECK WITH RAPHAEL).
These parameters were chosen to have the same spherical harmonic power spectrum for $\ell<1000$.
That means that it is not possible to distinguish between these cosmological models using these maps if the this range of scales is used.
We found that for $\ell>1000$ the differences in power spectrum is $~5\%$.
To remove this information, we additionally smooth the spherical maps with a Gaussian kernel of radius $3$ arcmins. The resulting PSD are displayed in Figure~\ref{fig:psd_sigma3}.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.95\textwidth]{figures/smooth_map_class_1.pdf}
\includegraphics[width=0.95\textwidth]{figures/smooth_map_class_2.pdf}
\caption{Example maps from two classes. Top: model 1 ($\Omega_m=0.31, \sigma_8=0.82$). Bottom: model 2 ($\Omega_m=0.26, \sigma_8=0.91$).
The initial conditions for these simulations were the same, so differences arise only due to different cosmological parameters. \nati{Tomek: Is this correct?}}
\label{fig:map_sample}
\end{figure*}

@@

Leveraging the properties of the HealPix sampling, we can split each maps
into $12*o^2$ ($o=1,2,4,\dots$) samples that span a smaller part of the sphere. 
The area of the pixels for an Nside of $o$ correspond to the different split of the data.
As shown in Figure~\ref{fig:subpart_sphere}, we used $o=1,2,4$ as the resulting
samples are large enough to suffer from the effects of the spherical geometry. We
decided to only report simulation results for this specific setting as spherical
cosmological data usually does not span the full sphere. Nevertheless, our code
includes an example using the full sphere.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/part_sphere.pdf}
\caption{3 subparts of the sphere with different sizes. Blue: $o=1$. Green: order $o=2$. Yellow: order $o=4$.}
\label{fig:subpart_sphere}
\end{figure}

The goal is build a robust classifier, i.e. an automatic way to
discriminate between the two classes in the presence of noise. For all the
experiments, we work with centered Gaussian noise. \todo{Tomek: could you put a
justification.} All classifiers are trained with noisy samples. To improve the
stability of the spherical CNN, we slowly increase the amount of noise added to
the training data over the iterations. Furthermore in order to avoid
overfitting, the noise is randomly produced during the training process. We trained our
model between $50$ and $200$ epochs depending on the complexity and the size of
the problem. The testing set was simply used to assess the global performance of
the network and to assess if it was overfitting.
Training took less that a day for each model using a Nvidia 1080 TI.

We compare the performance of the spherical CNN to two simple benchmark algorithms.
The two baselines are based on building features with a) the Power Spectral Densities or b) the histogram of pixels in the maps.
Those features are then classified use a Support Vector Machine (SVM) with a linear kernel. Other kernels did provide better result while having worse scaling properties.
For fairness, we augmented the dataset of the baseline in a similar way as for the network, i.e:

The histogram features contains information about the distribution of pixels, but does not include any spatial information.
Both the PSD features and the CNN can exploit the spatial information.

For The maps
The performance is assessed by measuring the classification error on a test set.

We compare the spherical CNN with a handcrafted feature classifier. CITE \cite{...}
The features are
constructed using a normalized histogram of the values. We then train a linear
SVN classifier. We tried different kernels, but they were not leading to better
results. For a fair comparison, we increase the amount of training data of the
SVN classifier until the testing error plateau to a fix value.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/psd_sigma3.pdf}
\caption{Power spectrum densities of the noiseless maps. The maps have be smoothed with a Gaussian kernel of radius $3$ arcmins in the preprocessing step.}
\label{fig:psd_sigma3}
\end{figure}

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/hist_error_order4_noise1_5.pdf}
\caption{TODO}
\label{fig:hist_error_evolution}
\end{figure}
